
\begin{acks}
{
%\footnotesize
%\fontsize{8.5}{9.5} \selectfont
\small
Many thanks to Paul Hovland (ANL) for plenty of thoughtful suggestions on automatic differentiation.
The authors would like to thank Lilly Chin (MIT) for their help in performing numerous hardware upgrades, without which this work would not have been possible.

% The information, data, or work presented herein was funded
% in part by NSF Grant OAC-1835443 and by DARPA under agreement number HR0011-20-9-0016 (PaPPa). 

% This material is based upon work supported by the Department of Energy, National Nuclear Security Administration under Award Number DE-NA0003965.

William S. Moses was supported in part by a DOE Computational Sciences Graduate Fellowship
DE-SC0019323. Valentin Churavy was supported in part by the Defense Advanced Research
Projects Agency (DARPA) under Agreement No. HR0011-20-9-0016, and in part by NSF Grant
OAC-1835443. Ludger Paehler was supported in part by the German Research Council (DFG) under grant agreement No. 326472365. This research was supported in part by Los Alamos National Laboratories grant
531711; in part by the Applied Mathematics activity within the U.S. Department of Energy, Office of Science, Advanced Scientific Computing Research Program, under contract number DE-AC02-06CH11357; in part by the Exascale Computing Project (17-SC-20-SC), a collaborative effort of two U.S. Department of Energy organizations (Office of Science and the National Nuclear Security Administration) responsible for the planning and preparation of a capable exascale ecosystem, including software,applications, hardware, advanced system engineering, and early testbed platforms, in support of the nation's exascale computing imperative; and in part by the United States Air Force Research Laboratory and the United States Air Force Artificial Intelligence Accelerator and was accomplished under Cooperative Agreement Number FA8750-19-2-1000. The views and conclusions contained in this document are those of the authors and should not be interpreted as representing the official policies, either expressed or implied, of the United States Air Force or the U.S. Government. The U.S. Government is authorized to reproduce and distribute reprints for Government purposes notwithstanding any copyright notation herein. We gratefully acknowledge the computing resources provided and operated by the Joint Laboratory for System Evaluation (JLSE) at Argonne National Laboratory.


% Disclaimer: "This report was prepared as an account of work sponsored by an agency of the United States Government. Neither the United States Government nor any agency thereof, nor any of their employees, makes any warranty, express or implied, or assumes any legal liability or responsibility for the accuracy, completeness, or usefulness of any information, apparatus, product, or process disclosed, or represents that its use would not infringe privately owned rights. Reference herein to any specific commercial product, process, or service by trade name, trademark, manufacturer, or otherwise does not necessarily constitute or imply its endorsement, recommendation, or favoring by the United States Government or any agency thereof. The views and opinions of authors expressed herein do not necessarily state or reflect those of the United States Government or any agency thereof."
}
\end{acks}
